# (PART) Scraping {-}

``` {r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(cache.path = 'cache/')
```

# First scraping with rvest {#rvest1}

With the knowledge of how an HTML file is constructed and how R and RStudio work
in basic terms, we are equipped with the necessary tools to take our first steps
in web scraping. In this session we will learn how to use the R package *rvest*
to read HTML source code into RStudio, extract targeted content we are 
interested in, and transfer the collected data into an R object for further 
analysis in the future.

## The rvest package

Part of the *tidyverse* is a package called rvest, which provides us with all
the basic functions for a variety of typical web scraping tasks. This package 
was included in the installation of the tidyverse package, but it is not part of
the *core tidyverse* and thus is not loaded into the current R session with
`library(tidyverse)`. Therefore, we have to do this explicitly:

``` {r, message = FALSE}
library(rvest)
```

The output in the RStudio console informs us that besides rvest, the package 
*xml2* has also been loaded, on which rvest is partly based and whose function
for reading HTML files, we will need in the following.

## hello_world.html

As a first exercise, it is a good idea to scrape the 
[Hello World](https://webscraping-tures.github.io/hello_world.html){target="_blank"}
example already described in chapter \@ref(html).
As a reminder, here is the HTML source code:

```
<!DOCTYPE html>

<html>
  <head>
    <title>Hello World!</title>
  </head>
  <body>
    <b>Hello World!</b>
  </body>
</html>
```

### read_html()

The first step in web scraping is to convert the page we are interested in into
an R object. This is made possible by the function `read_html()` from the xml2
package. `read_html()` "parses" the website, i.e. it reads the HTML, understands
its source code and transforms it into a representation R can understand. This
function needs the URL, i.e. the address of the website we want to read in, as 
its first argument. The URL must be given as a string, so we have to enclose it 
in `"`. The function also allows you to specify other options. In most cases, 
however, the default settings are sufficient. So we read in the hello_world.html 
file, assign it to a new R object at the same time and have this object put out
in the next step:

``` {r}
hello_world <- read_html("https://webscraping-tures.github.io/hello_world.html")
hello_world
```

As we can see in the output, the R object `hello_world` is a list with two 
entries. The first entry contains everything enclosed by the `<head>` tag, the
second entry everything enclosed by the `<body>` tag. The opening and closing
`<html>` tag is not part of the object. Remembering that HTML code is
hierarchically structured, the list is thus organised based on the highest 
remaining levels -- `<title>` and `<body>`.

We have thus successfully created a representation of the website in an R 
object. But what do we do with it now? In the case of this simple example, we 
might be interested in extracting the title of the website or the text displayed
on the page.

### html_nodes()

The function `html_nodes()` from the rvest package allows us to extract
individual elements of the HTML code. To do this, it needs the object to be
extracted from, as the first argument and a *selector* as well. In this 
introduction, we will concentrate exclusively on the so-called *CSS selectors*.
The alternative *XPath* is a bit more flexible, but CSS selectors are sufficient
in most cases and have a shorter and more intuitive syntax, which clearly makes
them the tool of choice here.

We will discuss the possibilities offered by CSS selectors in more detail in
chapter \@ref(css) and will limit ourselves to the basics for now. A selector in
the form `"tag"`, selects all HTML tags of the specified name. If we want to 
extract the `<title>` tag, we can do so in this way:

``` {r}
node_title <- html_nodes(hello_world, css = "title")
node_title
```

If we want to extract the text **Hello World!** shown on the website, one
possibility would be to select the complete `<body>` tag, since in this case no 
other text is displayed on the page.

``` {r}
node_body <- html_nodes(hello_world, css = "body")
node_body
```

This works in principle, but we also extracted the `<b>` tags as well as 
multiple *new lines* (`\n`), which we do not need both.
It would be more efficient to directly select the `<b>` tag enclosing the text.

``` {r}
node_b <- html_nodes(hello_world, css = "b")
node_b
```


### html_text()

In this case, we are interested in the text in the title and on the website, 
i.e. the content of the tags. We can extract this from the selected HTML 
elements in an additional step. This is made possible by the rvest function
`html_text()`. This requires the previously extracted HTML element as the only
argument.

``` {r}
html_text(node_title)
html_text(node_b)
```

With this, we have successfully completed our first web scraping goal, the 
extraction of the title and the text displayed on the page.

One more thing about the application of `html_text()` to elements that 
themselves contain further tags: Further above we extracted the object 
node_body, which contains the `<b>` tags as well as several line breaks in 
addition to the displayed text. Here, too, we can extract the pure text.

``` {r}
html_text(node_body)
```

We see that the function has conveniently removed the `<b>` tags we were not
interested in, for us. However, the line breaks and several spaces, so-called
*whitespace*, remain. Both can be removed with the additional argument 
`trim = TRUE`.

``` {r}
html_text(node_body, trim = TRUE)
```

## Countries of the World

Let us now look at a somewhat more realistic application. The website
<https://scrapethissite.com/pages/simple/>{target="_blank"}
lists the names of 250 countries, as well as their flag, capital, population and
size in square kilometres. Our goal could be to read this information into R for
each country so that we can potentially analyse it further.

Before we start, we should load the required packages (we will also need the 
tidyverse package this time) and read the website with the function 
`read_html()` and assign it to an R object.

``` {r, message = FALSE}
library(tidyverse)
library(rvest)

website <- read_html("https://scrapethissite.com/pages/simple/")
```

To understand the structure of the HTML file, the first step is to look at the
source code. As always, we can open it by right-clicking in the browser window 
and then clicking on "View Page Source". The first 100 or so lines of HTML code 
mainly contain information on the design of the website, which should not 
distract us further at this point. We are purely interested in the data of the
countries. The first country listed on the website is Andorra. It therefore 
makes sense to search the source code specifically for "Andorra". The key 
combination CTRL+F opens the search mask in your browser. We find what we are 
looking for in line 128. Since this source code, designed for practice purposes,
is formatted in a very structured way, we quickly realise that lines 125-135 are
code blocks related to Andorra. Let's look at these more closely:

```
<div class="col-md-4 country">
  <h3 class="country-name">
    <i class="flag-icon flag-icon-ad"></i>
    Andorra
  </h3>
  <div class="country-info">
    <strong>Capital:</strong> <span class="country-capital">Andorra la Vella</span><br>
    <strong>Population:</strong> <span class="country-population">84000</span><br>
    <strong>Area (km<sup>2</sup>):</strong> <span class="country-area">468.0</span><br>
  </div>
</div><!--.col-->
```

All the information about Andorra is enclosed in a `<div>` tag. As a reminder, a
`<div>` defines a grouping of code across multiple lines. In web design 
practice, these groupings are mainly used to assign a certain CSS style to the
following code via the argument `class=`, for example to define the typeface. 
From a web scraping perspective, we generally don't care how the styles are 
defined. We just need to know that we can exploit these CSS assignments of 
classes for our purposes. At the next level down, we find two blocks, one 
containing, among other things, the name of the country and another containing
information about that country. Let's look at the first block first.

### Country names

```
<h3 class="country-name">
  <i class="flag-icon flag-icon-ad"></i>
  Andorra
</h3>
```

The name "Andorra" is enclosed in an `<h3>` tag, i.e., a third-level heading. In
addition to the name, we also find another tag within the tag that includes the 
image of the flag. Since we are not interested in the graphics here, we can 
ignore this.

On this website, all `<h3>` tags are used exclusively to display the names of 
the countries. Thus, we can use the `<h3>` tag as a CSS selector to read out the
enclosed text analogous to the first example.

``` {r}
node_country <- html_nodes(website, css = "h3")
text_country <- html_text(node_country, trim = TRUE)

head(text_country, n = 10)
```

The result looks promising. Since the structure of the code block is the same 
for each country, the vector `text_country` was created in this way with 250 
entries, exactly the number of countries listed on the website. For reasons of
clarity, it often makes sense not to put out the complete and often very long 
vectors, data frames or tibbles, but to use the function `head()` to list the
number of entries specified by the argument `n`, starting with the first.

#### The pipe %>% 

At this point, we should think again about the readability and structure of our
R code. Let us consider the preceding code block:

``` {r, eval = FALSE}
node_country <- html_nodes(website, css = "h3")
text_country <- html_text(node_country, trim = TRUE)
```

As we have seen, this achieves our goal. However, we have also created the
`node_country` object to temporarily save the first step -- reading the `<h3>`
tags. We will never need this object again. If we use the already introduced 
pipe `%>%` instead, the need to cache partial results is eliminated and we write
code that is more intuitive and easier to understand at the same time.

``` {r}
country <- website %>% 
  html_nodes(css = "h3") %>% 
  html_text(trim = TRUE)

head(country, n = 10)
```

As a reminder, the pipe passes the result of a work step along to the next 
function, which in the tidyverse as well as in many other R-functions (but not 
ALL!) takes data as the first argument, which we then do not have to define
explicitly. For a better understanding, let's look at the above example in 
detail. The first line passes the object `website` along to the function 
`html_nodes()`. So we don't have to tell `html_nodes()` which object to apply 
to, because we already passed it along to the function with the pipe. The 
function is applied to the object `website` with all other defined arguments -- 
here `css` -- and the result is passed along again to the next line, where the
`html_text()` function is applied to it. Here the pipe ends, and the final 
result is assigned to the object country.

We now need three instead of two lines to get the same result, but the actual 
typing work has been reduced -- especially if you create the pipe with the key
combination CTRL+Shift+M -- and we have created code that can be read and
understood more intuitively with a little practice.

So should we always connect all steps with the pipe? No. In many cases it makes 
sense to save intermediate results in an object, namely whenever we will access 
it multiple times. In our example, we could also integrate the import of the 
website into the pipe:

``` {r}
country <- read_html("https://scrapethissite.com/pages/simple/") %>% 
  html_nodes(css = "h3") %>% 
  html_text(trim = TRUE)
```

Overall, this saves us even more typing. However, since we still have to access 
the selected website multiple times later on, this would also mean that the 
parsing process has to be repeated each time. On the one hand, this can have a
noticeable impact on the computing time for larger amounts of data. On the other
hand, it also means accessing the website's servers and downloading the data 
again each time. However, we should avoid data traffic generated without good 
reasons as part of a *good practice* of web scraping. So it makes perfect sense
to save the result of the `read_html()` function in an R object so that it can 
be reused multiple times later.

### Capitals, population and area

Let us now turn to the further information for each country. These are located 
in the second block of the HTML code considered above:

```
<div class="country-info">
  <strong>Capital:</strong> <span class="country-capital">Andorra la Vella</span><br>
  <strong>Population:</strong> <span class="country-population">84000</span><br>
  <strong>Area (km<sup>2</sup>):</strong> <span class="country-area">468.0</span><br>
</div>
```

As we can see, both the name of the capital, the population of the country, and 
its size in square kilometers are enclosed by a `<span>` tag in lines 2--4
respectively. Like `<div>`, `<span>` defines groupings, but not across multiple 
lines but for one, or as here, part of a line. So let's try to read the names of
the capitals, using the `<span>` tag as a selector.

``` {r}
website %>% 
  html_nodes(css = "span") %>% 
  html_text() %>% 
  head(n = 10)
```

So we get the names of the capitals, but also the population and the size of the
country. `span` was too unspecific as a selector. Since all three types of
country data are enclosed with `<span>` tags, all three are also selected. So we
have to tell `html_nodes()` more precisely which `<span>` we are interested in. 
This is where the CSS classes we mentioned earlier come into play. These differ
between the three countries’ information. For example, the `<span>` that 
includes the name of the capital city is assigned the class `"country-capital"`.
We can target this class with our CSS selector. To select a class, we can use 
the syntax `.class-name`. So, to select all `<span>` that have the class
`"country-capital"`, we can do as follows:

``` {r}
capital <- website %>% 
  html_nodes(css = "span.country-capital") %>% 
  html_text()

  head(capital, n = 10)
```

We can repeat this in an analogue manner for the number of inhabitants with the 
class `"country-population"`.

``` {r}
population <- website %>% 
  html_nodes(css = "span.country-population") %>% 
  html_text()

  head(population, n = 10)
```

If we take a closer look at the vector created in this way, we see that it is a
character vector. For inspection we can use the function `str()`, which gives us
the structure of an R object, including the data type used.

``` {r}
str(population)
```

So the numbers were not read out as numbers but as strings. Among other things,
this does not allow for calculation with the numbers. `population[1]` selects 
the first element of the vector.

``` {r, error = TRUE}
population[1] / 2
```

As you remember, one way to tell R to interpret the "text" read from the HTML
code as numbers is to use the function `as.numeric()`.

``` {r}
population <- website %>% 
  html_nodes(css = "span.country-population") %>% 
  html_text() %>% 
  as.numeric()

str(population)

population[1] / 2
```

In the same way, the size in square kilometers can be read with the class
`"country-area"`.

``` {r}
area <- website %>% 
  html_nodes(css = "span.country-area") %>% 
  html_text() %>% 
  as.numeric()

str(area)
```


### Merge into one tibble

We have now created four vectors, which respectively contain the information 
about the name of the country, the associated capital, the number of population 
and the size of the country. For Andorra:

``` {r}
country[1]
capital[1]
population[1]
area[1]
```

We could already continue working with this, but for many applications it is 
more practical if we compile the data in tabular form. In the tidyverse, the 
form of the tibble is suitable for this purpose.

``` {r}
countries <- tibble(
  Land = country,
  Hauptstadt = capital,
  Bevoelkerung = population,
  Flaeche = area
)

countries
```

This is not only more readable but also facilitates all further potential 
analysis steps.

If we are sure that we do not need the individual vectors, we can also perform
the reading of the data and the creation of the tibble in a single step. Below 
you can see how the complete scraping process can be completed in relatively few
lines.

``` {r}
website <- "https://scrapethissite.com/pages/simple/" %>%
  read_html()

countries_2 <- tibble(
  Land = website %>%
    html_nodes(css = "h3") %>% 
    html_text(trim = TRUE),
  Hauptstadt = website %>% 
    html_nodes(css = "span.country-capital") %>% 
    html_text(),
  Bevoelkerung = website %>% 
    html_nodes(css = "span.country-population") %>% 
    html_text() %>% 
    as.numeric(),
  Flaeche = website %>% 
    html_nodes(css = "span.country-area") %>% 
    html_text() %>% 
    as.numeric()
)

countries_2
```