# Scraping of tables & dynamic websites {#rvest2}

``` {r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(cache.path = 'cache/')
```

## Scraping of tables

In web scraping, we will often pursue the goal of transferring the extracted 
data into a tibble or data frame in order to be able to analyse it further. It
is particularly helpful if the data we are interested in is already stored in an
HTML table. Because rvest allows us to read out complete tables quickly and
easily with the function `html_table()`.


As a reminder, the basic structure of the HTML code for tables is as follows:

```
<table>
  <tr> <th>#</th> <th>Tag</th> <th>Effect</th> </tr>
  <tr> <td>1</td> <td>"b"</td> <td>bold</td> </tr>
  <tr> <td>2</td> <td>"i"</td> <td>italics</td> </tr>
</table>
```

The `<table>` tag covers the entire table. Rows are defined by `<tr>`, column
headings with `<th>` and cells with `<td>`.

Before we start scraping, we load the necessary packages as usual:

``` {r, message = FALSE}
library(tidyverse)
library(rvest)
```

### Table with CSS selectors from Wikipedia

On the
[Wikipedia page on "CSS"](https://en.wikipedia.org/wiki/CSS){target="_blank"},
there is also a table with CSS selectors. This is our scraping target.

First we parse the website:

``` {r}
website <- "https://en.wikipedia.org/wiki/CSS" %>% 
  read_html()
```

If we look at the source code and search -- CTRL+F -- for "<table", we see that 
this page contains a large number of HTML tables. These include not only the
elements that are recognisable at first glance as "classic" tables, but also, 
among other things, the "info boxes" at the top right edge of the article or the
fold-out lists of further links at the bottom. If you want to look at this more
closely, the Web Developer Tools can be very helpful here.

Instead of simply selecting all `<table>` nodes on the page, one strategy might
be to use the WDTs to create a CSS selector for that specific table:
`"table.wikitable:nth-child(28)"`. We thus select the table of class 
`"wikitable"` which is the 28th child of the parent hierarchy level -- 
`<div class="mw-parser-output">`.

If we only want to select a single HTML element, it can be helpful to use the 
function `html_node()` instead of `html_nodes()`.

``` {r}
node <- website %>% 
  html_node(css = "table.wikitable:nth-child(28)")
nodes <- website %>% 
  html_nodes(css = "table.wikitable:nth-child(28)")

nodes
node
```

The difference is mainly in the output of the function. This is recognisable by 
the entry inside the { } in the output. In the first case, we get a list of HTML 
elements -- an "xml_nodeset" -- even if this list, as here, consists of only one
entry. `html_node()` returns the HTML element itself -- the "html_node" -- as
the function's output. Why is this relevant? In many cases it can be easier to 
work directly with the HTML element instead of a list of HTML elements, for 
example when transferring tables into data frames and tibbles, but more on that 
later. 

To read out the table selected in this way, we only need to apply the function 
`html_table()` to the HTML node.

``` {r}
css_table_df <- node %>% 
  html_table()

css_table_df %>% 
  head(n = 4)
```

The result is a data frame that contains the scraped contents of the HTML table
and adopts the column names stored in the `<th>` tags for the columns of the 
data frame. Due to the very long cells in the "Matches" column, the output of 
the RStudio console is unfortunately not particularly helpful. Another advantage
of using tibbles instead of data frames is that long cell contents are
automatically abbreviated in the output. To convert a data frame into a tibble, 
we can use the function `as_tibble()`.

``` {r}
css_table_tbl <- node %>% 
  html_table() %>% 
  as_tibble()

css_table_tbl %>% 
  head(n = 4)
```


### Scraping multiple tables

It could also be our scraping goal to scrape not only the first, but all four
content tables of the Wikipedia article. If we look at the four tables in the
source code and/or the WDTs, we see that they all carry the class `"wikitable"`.
This allows us to select them easily. Please note that the function 
`html_nodes()` must be used again, as we no longer need just one element, but a
list of several selected elements.

``` {r}
tables <- website %>% 
  html_nodes(css = "table.wikitable") %>% 
  html_table()
```

The result is a list of four data frames, each of which contains one of the four
tables. If we want to select an individual data frame from the list, for 
example, to transfer it into a new object, we have to rely on *subsetting*.

We have learned about basic subsetting for vectors using `[#]`, in chapter
\@ref(R1). For lists, things can get a little bit more complicated.
There are basically two ways of subsetting lists in R: `list_name[#]` and 
`list_name[[#]]`. The most relevant difference for us is what kind of object R
returns to us. In the first case, the returned object is always a list, even if
it may only consist of one element. Using double square brackets, on the other 
hand, returns a single element directly. So the difference is not dissimilar to
that between `html_nodes()` and `html_node()`.

For example, if our goal is to select the third data frame from the list of four 
data frames, which subsetting should we use?

``` {r}
tables[3] %>% 
  str()

tables[[3]] %>% 
  str()
```

In the first case, we see that we have a list of length 1, which contains a data 
frame with 7 lines and 2 variables, as well as further information about these
variables. In the second case, we get the data frame directly, i.e. no longer as
an element of a list. So we have to use `list_name[[]]` to directly select a 
single data frame from a list of data frames.

If we are interested in selecting several elements from a list instead, this is
only possible with `list_name[]`. Instead of selecting an element with a single
number, we can select several with a vector of numbers in one step.


``` {r}
tables[c(1, 3)] %>% 
  str()
```

As a result, we get a list again that contains the two elements selected here.


### Tabellen mit NAs

What happens when we try to read a table with missing values? Consider the 
following example:
<https://webscraping-tures.github.io/table_na.html>{target="_blank"}

At first glance, it is already obvious that several cells of the table are
unoccupied here. Values are missing. Let's try to read in the table anyway.

``` {r error = TRUE}
table_na <- "https://webscraping-tures.github.io/table_na.html" %>% 
  read_html %>% 
  html_node(css = "table")

table_na %>% 
  html_table()
```

We get a helpful error message informing us that the number of columns is not 
constant across the entire table. We are also kindly offered a possible solution
directly. The function `html_table()` can be instructed with the argument 
`fill = TRUE` to automatically fill rows with a different number of columns with
`NA`. This stands for "Not Available" and represents missing values in R.

``` {r}
wahlbet <- table_na %>% 
  html_table(fill = TRUE)

wahlbet %>% 
  head(n = 4)
```

As we can see, `html_table` was able to fill the four cells with missing values
with `NA` and read the table despite the problems. However, there are two
different types of problems in the HTML source code, which the automatic repair 
handles differently. Let's first look at the source code of the first two lines:

```
<tr>
  <td>Baden-Württemberg</td>
  <td>2016</td>
  <td></td>
</tr>

<tr>
  <td>Bayern</td>
  <td>2018</td>
</tr>
```

For "Baden-Württemberg", we see that the third column is created in the source 
code, but there is no content in this cell. `html_table()` could have read this
without `fill = TRUE` and would have filled the cell with a `NA`. In contrast, 
for "Bayern" the cell is completely missing. This means that the second row of
the table consists of only two columns, while the rest of the table has three
columns. This is the problem that the error message pointed out to us. As a 
result, R was able to draw the correct conclusion in both cases and fill the 
cell in both rows with an `NA`.

But let's also look at the third and fourth rows in the source code:

```
<tr>
  <td>Berlin</td>
  <td></td>
  <td>66.9</td>
</tr>

<tr>
  <td>Brandenburg</td>
  <td>61.3</td>
</tr>
```

The second column is missing in both cases. In the first case it is created but
empty, in the second it does not exist. In the first case, `html_table()` can
again handle it without any problems. For "Brandenburg", however, the function
reaches its limits. We, as human observers, quickly realise that the last state
election in Brandenburg did not take place in 61.3 and that this must therefore
be the turnout. R cannot distinguish this so easily and takes 61.3 as the value
for the column "Election year" and inserts a `NA` for "Voter turnout".

What to do? First of all, we should be aware that such problems exist. So if 
`html_table()` gives this error message, we should not simply set `fill = TRUE`,
but try to find out why the problem exists and whether the option to have it
fixed automatically will actually get us there. If this is not the case, one
approach could be to write our own *extractor function* that fixes the problems 
directly during scraping. However, this is a rather advanced method and outside
the scope of what is feasible in this introduction. However, we can at least 
correct the problems that arise afterwards.

Our problem lies exclusively in row four. Its second column must be moved to the
third and the second must then itself be set as `NA`. For this we need 
subsetting again. In the case of a data frame, we need to specify the row and 
column in the form `df[row, column]` to select a cell. So we can tell R: "Write
in cell three the content of cell two, and then write in cell two `NA`".

```{r}
wahlbet[4, 3] <- wahlbet[4, 2]
wahlbet[4, 2] <- NA

wahlbet %>% 
  head(n = 4)
```

## Dynamische Websites

In the "reality" of the modern internet, we will increasingly encounter websites 
that are no longer based exclusively on static HTML files, but generate content 
dynamically. You know this, for example, in the form of timelines in social 
media offerings that are generated dynamically based on your user profile. Other
websites may generate the displayed content with *JavaScript* functions or in
response to input in HTML forms.

In many of these cases, it is no longer sufficient from a web scraping 
perspective to read in an HTML page and extract the data you are looking for, as
this is often not contained in the HTML source code but is loaded dynamically in
the background. The good news is that there are usually ways of scraping the
information anyway.

Perhaps the operator of a page or service offers an *API* (Application 
Programming Interface). In this case, we can register for access to this 
interface and then get access to the data of interest. This is possible with
Twitter, for example. In other cases, we may be able to identify in the embedded
scripts how and from which database the information is loaded and access it
directly. Or we use the *Selenium WebDriver* to "remotely control" a browser 
window and scrape what the browser "sees".

However, all of these approaches are advanced methods that are beyond the scope 
of this introduction.

But in cases where an HTML file is dynamically generated based on input into a
HTML form, we can read it using the methods we already know.

### HTML forms and HTML queries

As an example, let's first look at the OPAC catalogue of the Potsdam University
Library <https://opac.ub.uni-potsdam.de/>{target="_blank"} in the browser.

If we enter the term "test" in the search field and click on Search, the browser
window will show us the results of the search query. But what actually interests 
us here is the browser's address bar. Instead of the URL 
"https://opac.ub.uni-potsdam.de/", there is now a much longer URL in the form: "https://opac.ub.uni-potsdam.de/DB=1/CMD?ACT=SRCHA&IKT=1016&SRT=YOP&TRM=test". 
The first part is obviously still the URL of the website called up, let's call
this the base URL. However, the part "CMD?ACT=SRCHA&IKT=1016&SRT=YOP&TRM=test" 
was added to the end of the URL. This is the *HTML query* we are interested in
here. Between the base URL and the query there are one or more components, which
in this case may also differ depending on your browser. However, these are also
irrelevant for the actual search query. The shortened URL: "https://opac.ub.uni-potsdam.de/CMD?ACT=SRCHA&IKT=1016&SRT=YOP&TRM=test" 
leads to the same result.

A query is a request in which data from an HTML form is sent to the server. In 
response, the server generates a new website, which is sent back to the user and
displayed in the browser. In this case, the query was triggered by clicking on 
the "Search" button. If we understand what the components of the query do, we
could manipulate it and use it specifically to have a website of interest 
created and parsed.


### HTML forms

To do this, we first need to take a look at the HTML code of the search form. To
understand this, you should display the source code of the page and search for 
"<form" or use the WDTs to look at the form and its components.

```
<form action="CMD"
      class="form"
      name="SearchForm"
      method="GET">
  ...
</form>
```

HTML forms are encompassed by the `<form>` tag. Within the tag, one or more form
elements such as text entry fields, drop-down option lists, buttons, etc. can be 
placed.

`<form>` itself carries a number of attributes in this example. The first
attribute of interest to us is the `method="GET"` attribute. This specifies the
method of data transfer between client and server. It is important to note that 
the method "GET" uses queries in the URL for the transmission of data and the
method "POST" does not. We can therefore only manipulate queries if the "GET" 
method is used. If no method is specified in the `<form>` tag, "GET" is also
used as the default.

The second attribute of interest to us is `action="CMD"`. This specifies which
action should be triggered after the form has been submitted. Often the value of
`action=` is the name of a file on the server to which the data will be sent and
which then returns a dynamically generated HTML page back to the user.

Let us now look at the elements of the form. For this, the rvest function 
`html_form()` can be helpful.

``` {r}
"https://opac.ub.uni-potsdam.de/" %>% 
  read_html() %>% 
  html_node(css = "form") %>% 
  html_form()
```

The output shows us in the first line again the values for `method=` and 
`action=` as well as the name of the form. The other six lines show that the 
form consists of three `<select>` and three `<input>` elements. We also see the
names of these elements as well as the default value that is sent when the form
is submitted, as long as no other value is selected or entered.

Let's look at some of these elements. `<select>` elements are drop-down lists of
options that can be selected. This is the source code for the first `<select>`
element in our example:

```
<select name="ACT">
  <OPTION VALUE="SRCH">suchen [oder]
  <OPTION VALUE="SRCHA" SELECTED>suchen [und]
  <OPTION value="BRWS">Index bl&auml;ttern
</select>
```

The attribute `name="ACT"` defines the elements name, which is used when 
transmitting the data from the form via the query. The `<option>` tags define 
the selectable options, i.e. the drop down menu. `<value="">` represents the 
value transmitted by the form. The user is being shown the text following the 
tag. The default selection is either the first value in the list or -- like in
this case -- the option with the attribute `selected` is being explicitly chosen
as the default.

The three other elements are `<input>` tags. Input fields whose specific type is
specified via the attribute `type=""`. These can be, for example, text boxes
(`type="text"`) or checkboxes (`input="checkbox"`), but there are many more
options available. A comprehensive list can be found at:
<https://www.w3schools.com/html/html_form_input_types.asp>{target="_blank"}. 
Here is the source code for two of the three `<input>` elements on the example
page:

```
<input type="text" name="TRM" value="" size="50">

<input type="submit" class="button" value=" Suchen ">
```

The first tag is of the type "text", i.e. a text field, in this case the text
field into which the search term is entered. In addition to the name of the
element, a default value of the field is specified via `value=""`. In this case,
the default value is an empty field. The second tag is of the type "submit". 
This is the "Search" button, which triggers the transmission of the form data 
via the query by clicking on it.


### The query

But what exactly is being transmitted? Let's look again at the example query 
from above:

```
CMD?ACT=SRCHA&IKT=1016&SRT=YOP&TRM=test
```

The value of the `action=""` attribute forms the first part of the query and is
appended after the base URL. The value of the attribute tells the server what to
do with the other transmitted data. This is followed by a `?`, which introduces 
the data to be transmitted as several pairs of `name=""` and `value=""` 
attributes of the individual elements. The pairs are connected with `&`. 
`ACT=SRCHA` thus stands for the fact that the option "SRCHA" has been selected 
in the element with the name "ACT". What the values of the two other `<select>`
elements "IKT" and "SRT" stand for, you can understand yourself with a look into
the source code or the WDTs. The text entered in the field is transmitted as the
value of the `<input type="text">` with the name "TRM". Here "test".

The server receives the form data in this way, can then take a decision on the 
basis of the `action=""` attribute, here "CMD", how the data is to be processed 
and constructs the website accordingly, which it sends back to us and which is 
displayed in our browser.


### Manipulating the query and scraping the result

Now that we know what the components of the query mean, we can manipulate them. 
Instead of writing queries by hand, we should use R to combine them for us. We
will also encounter the technique of manipulating URLs directly in the R code
more often. So we should learn it early.

The function `str_c()` assembles the strings listed as arguments, i.e. sequences
of letters, into a single string. Strings stored in other R objects can also be
included. If we have the goal of manipulating both the search method and the
search handle, we could achieve this in this way:

``` {r}
base_url <- "https://opac.ub.uni-potsdam.de/"
method <- "SRCHA"
term <- "test"

url <- str_c(base_url, "CMD?ACT=", method, "&IKT=1016&SRT=YOP&TRM=", term)
url
```

If we now change the strings stored in the `method` and `term` objects and
generate the complete URL again, these components of the query are manipulated
accordingly.

``` {r}
method <- "SRCH"
term <- "web+scraping"

url <- str_c(base_url, "CMD?ACT=", method, "&IKT=1016&SRT=YOP&TRM=", term)
url
```

The search method was set to the value "SRCH", i.e. an "OR" search, the search 
term to "web scraping". It is important to note that no spaces may appear in the
query and that these are replaced by "+" when the form is submitted. So instead
of "web scraping" we have to use the string "web+scraping".

As an example application, we can now have the server perform an "AND" search 
for the term "web scraping", read out the HTML page generated by the server and
extract the 10 titles displayed.

``` {r}
base_url <- "https://opac.ub.uni-potsdam.de/"
method <- "SRCHA"
term <- "web+scraping"

url <- str_c(base_url, "CMD?ACT=", method, "&IKT=1016&SRT=YOP&TRM=", term)
url

website <- url %>% 
  read_html()
```

The search results are displayed as tables in the generated HTML file. The
`<table>` tag has the attribute-value combination `summary="hitlist"`, which we
can use for our CSS selector:

``` {r}
hits <- website %>% 
  html_node(css = "table[summary='hitlist']") %>% 
  html_table() %>% 
  as_tibble()

hits %>% head(n=10)
```

This worked, but we see that the table consists mainly of empty rows and cells. 
These are invisible on the website, but are used to format the display. 
Instead of repairing the table afterwards, it makes more sense to extract only
the cells that contain the information we are looking for. These are the `<td>`
tags with `class="hit"` and the attribute-value combination `align="left"`. 
On this basis, we can construct a unique CSS selector.

``` {r}
hits <- website %>% 
  html_nodes(css = "td.hit[align='left']") %>% 
  html_text(trim = TRUE)

hits %>% head(n = 5)
```

### Additional resources

In order to process this information further and, for example, separate it into 
data on author, title, year, etc., advanced knowledge in dealing with strings is
necessary, which unfortunately goes beyond the scope of this introduction. A 
good first overview can be found in the chapter "Strings" from "R for Data
Science" by Wickham and Grolemund: 
<https://r4ds.had.co.nz/strings.html>{target="_blank"}

The appropriate "cheat sheet" is also recommended:
<https://raw.githubusercontent.com/rstudio/cheatsheets/master/strings.pdf>{target="_blank"}