# (PART) Data analysis {-}

# Scraping Berlin police reports {#scraping}

``` {r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(cache.path = 'cache/')
```


From here on out, we will talk about the basics of data analysis, and of using
regular expressions when dealing with textual data. We will start by showing,
and briefly explaining, how we scrape the data we will be using as an example
for the remaining chapters.

The website
<https://www.berlin.de/polizei/polizeimeldungen/archiv/>{target="_blank"}
contains reports by the Berlin police that are open to the public, beginning
with the year 2014. We will gather the links to all subpages, download them and
extract the text of the report, as well as the date, time and district where it
occurred.

## Gathering the links

``` {r message = FALSE}
library(tidyverse)
library(rvest)
```

We begin by downloading the mainpage.

``` {r}
website <- "https://www.berlin.de/polizei/polizeimeldungen/archiv/" %>% 
  read_html()
```

The next step is to extract the URLs for the yearly archives. All the `<a>` tags
that contain those links, have a `title=` attribute whose value begins with
"Link". We can use this in selector construction, extract the value of the
`href=` attribute and, as they are incomplete URLs, append them to the base URL.

``` {r}
links <- website %>% 
  html_nodes(css = "div.html5-section.body a[title^='Link']") %>% 
  html_attr(name = "href") %>% 
  str_c("https://www.berlin.de", .)
```

Each yearly archive page contains several subpages for the reports in that year,
divided by pagination; to gather the links to all the subpages, we have to
understand how the pagination works. In this case, the URL for the yearly
archive is simply appended with the query "?page_at_1_0=" and a value indicating
the number of the subpage we want to access.
So "https://www.berlin.de/polizei/polizeimeldungen/archiv/2020/?page_at_1_0=5"
for example gives access to the fifth page for 2020. 

The number of police reports is not constant over the years; the number of
subpages per year will also differ. We do not know in advance, how many
subpages we will have to download, but we can scrape this information from the
last pagination link for each year. We read the first page for each year, select
the `<a>` tag for the last subpage, extract the text from it and save it as a
number. Note that I also added a waiting time of two seconds between each
download, as shown in chapter \@ref(goodpractice). 

``` {r}
max_pages <- links %>% 
  map(~ {
    Sys.sleep(2)
    read_html(.x)
  }) %>%
  map(html_node, css = "li.pager-item.last > a") %>% 
  map(html_text) %>% 
  as.numeric()
```

Now we can construct the links to all subpages; done by nesting two for loops in
my script.

```{r}
pag_links <- character()

for (i in 1:length(links)) {
  for (j in 1:max_pages[i]) {
    pag_links <- append(pag_links, values = str_c(links[i], "?page_at_1_0=", j))
  }
}
```


The outer loop is a counter for the elements in the objects `links`; in this
case the loop counts from 1 to 8, one step for each yearly archive page. For
each of these iterations, the inner loop will count from 1 to the number of
the last subpage for this year, which we assigned to the vector `max_pages`.
Using both counters together, we can construct the links to the subpages using
`str_c()`. The function takes the link to a yearly archive indicated by the
counter in the outer loop `i`, appends "?page_at_1_0=" and after this the value
of the inner loop counter `j`. The resulting link is then added to the vector
`pag_links` using the function `append()` which adds the data indicated by the
argument `values` to the object specified in the first argument. For this to
work, we first have to initialise the object as an empty object outside of the
loop. Here I created it as an empty character vector, using
`pag_links <- character()`. If you want to create an empty object without any
data type assigned, you can also use `empty_vector <- NULL`.


## Downloading the subpages

Now that the the links to all subpages are gathered, we can finally download
them. Please note that the download will take up to 20 minutes due to the number
of subpages (337 at the time of writing), and the additional waiting time of 2
seconds between each iteration.

``` {r}
pages <- pag_links %>% 
  map(~ {
    Sys.sleep(2)
    read_html(.x)
  })
```

## Extracting the data of interest

The goal is to extract the text of the report, the date/time and the district
the report refers to. Let us try to approach this using the techniques we
developed over the last chapters:

``` {r error=TRUE}
reports <- tibble(
  Date = pages %>% 
    map(html_nodes, css = "div.cell.date") %>% 
    map(html_text) %>% 
    unlist(),
  Report = pages %>% 
    map(html_nodes, css = "div.cell.text > a") %>% 
    map(html_text) %>% 
    unlist(),
  District = pages %>% 
    map(html_nodes, css = "span.category") %>% 
    map(html_text) %>% 
    unlist()
)
```

That did not work. But why? Let us look at the error message we received. It
informs us that the column "District" we tried to create, is shorter in length
compared to the other two. Since we can not create tibbles out of columns with
different lengths, we get this error.

The fact that the "District" column is shorter than the other two, must mean
that there are police reports where the information on the district is not
listed on the website, which we can confirm by browsing some of the subpages,
e.g.
<https://www.berlin.de/polizei/polizeimeldungen/archiv/2014/>{target="_blank"}.


### Dealing with missing data in lists

Some of the `<span>` tags that contain the district are missing. Using the
approach presented above, `html_nodes()` just extracts the `<span>` tags that
are present. We tell R that we do want all `<span>` tags of the class
"category", and this is what R returns to us. The cases where the tag is
missing, are skipped. But this is not what we want; what we actually want is,
that R looks at every single police report and saves its text, date and time, as
well as the district, if it is not missing. If it is missing, we want R to save
a `NA` in the cell of the tibble, the representation of missing values in R.

The `<div>` and `<span>` tags that contain the data of interest are nested in
`<li>` tags in this case. The `<li>` tags thus contain the whole set of data,
the text, the date/time and the district. The approach here is to make R examine
every single `<li>` tag and extract the data that is present, as well as save an
`NA` for every piece of data that is missing.

To start, we have to extract all the `<li>` tags and their content from the
subpages. Right now, the subpages are saved in the object `pages`. We use a for
loop that takes every element of `pages`, extracts all the `<li>` tags from it
and adds them to a new list using `append()`. `append()` takes an existing
object as its first argument and adds the data indicated in the `values =`
argument to it. For this to work, we have to initiate the new list as an empty
object before the loop starts.

``` {r}
list_items <- NULL

for (i in 1:length(pages)) {
  list_items <- append(list_items, values = html_nodes(pages[[i]], css = "li.row-fluid"))
}
```

The newly created list `list_items` contains a nodeset for each `<li>` tag from
all subpages:

``` {r}
list_items[[1]]
```

We can now make R examine every element of this list one after the other and
extract all the data they contain. But what happens when we try to extract a
node that is not present in one of the elements? R returns an `NA`:

``` {r}
html_node(list_items[[1]], css = "span.notthere")
```

Using a for loop that iterates over the elements in `list_items`, we write a
tibble row by row, filling the data cells with the extracted information and
with `NA` if a node could not be found for this element of `list_items`. We have
to initiate the tibble before the for loop starts: we define the column names,
the type of data to be saved and also the length of the columns. The
latter is not strictly necessary as we could also have created a tibble with a
column length of $0$, but pre-defining their length increases computational
efficiency. Still, the for loop has to iterate over several thousand elements
and extract the data contained, which will take several minutes to complete.

``` {r}
reports <- tibble(
  Date = character()[1:length(list_items)],
  Report = character()[1:length(list_items)],
  District = character()[1:length(list_items)]
)

for (i in 1:length(list_items)) {
  reports[i, "Date"] <- html_node(list_items[[i]], css = "div.cell.date") %>% 
    html_text()
  reports[i, "Report"] <- html_node(list_items[[i]], css = "div.cell.text > a") %>% 
    html_text()
  reports[i, "District"] <- html_node(list_items[[i]], css = "span.category") %>% 
    html_text()
}
```


## Saving the data

As discussed in chapter \@ref(files), we save the scraped data at this point.
You have seen that we downloaded a lot of subpages, which took a considerable
amount of time; if we repeat this for every instance of further data analysis,
we create a lot of unnecessary traffic and waste a lot of our own time. 

``` {r}
save(reports, file = "reports.RData")
```
