# Graphical analysis with ggplot {#ggplot}


``` {r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(cache.path = 'cache/')

knitr::opts_chunk$set(eval = FALSE)
```

WIP! Content is due to change!


## ggplot syntax

## Aesthtetics

### Continuous

### Continuous x Continuous

### Discrete x Continuous

## Tuning the output





``` {r}
ggplot(data = data) +
  geom_point(aes(x = eligible, y = turnout)) +
  geom_line(aes(x = eligible, y = turnout)) +
  geom_smooth(aes(x = eligible, y = turnout))
```



``` {r}
data %>%
  group_by(gender) %>% 
  summarise(
    eligible_total = sum(eligible),
    turnout_total = sum(turnout),
    turnout_pct_mean = weighted.mean(turnout_pct, w = eligible)
  ) %>% 
  ggplot() +
  geom_col(aes(x = gender, y = turnout_pct_mean))
```

``` {r}
data %>%
  group_by(ybirth) %>% 
  summarise(
    eligible_total = sum(eligible),
    turnout_total = sum(turnout),
    turnout_pct_mean = weighted.mean(turnout_pct, w = eligible)
  ) %>% 
  ggplot() +
  geom_col(aes(x = ybirth, y = turnout_pct_mean))
```

``` {r}
overall_turnout_mean <- data %>% 
  summarise(
    turnout_pct_mean = weighted.mean(turnout_pct, w = eligible)
  ) %>% 
  unlist()

data %>%
  group_by(ybirth) %>% 
  summarise(
    turnout_pct_mean = weighted.mean(turnout_pct, w = eligible)
  ) %>% 
  ggplot() +
  geom_col(aes(x = ybirth, y = turnout_pct_mean)) +
  geom_hline(yintercept = overall_turnout_mean)
```

turnout_pct_mean = 
  
  
  ``` {r}
data %>%
  ggplot() +
  geom_col(aes(x = ybirth, y = turnout_pct, fill = gender), position = "dodge") +
  geom_hline(yintercept = overall_turnout_mean)
```



``` {r}
data %>%
  ggplot() +
  geom_col(aes(x = ybirth, y = turnout_pct, fill = gender, colour = eligible), position = "dodge") +
  geom_hline(yintercept = overall_turnout_mean) +
  geom_text(aes(x = ybirth, y = turnout_pct, fill = gender, label = eligible), position = position_dodge(width = 1), hjust = -2, angle = -90)
```


``` {r}
data %>%
  ggplot(aes(x = ybirth, y = turnout_pct, fill = gender)) +
  geom_col(aes(colour = eligible), position = "dodge") +
  geom_hline(yintercept = overall_turnout_mean) +
  geom_text(aes(label = eligible), position = position_dodge(width = 1), hjust = -2, angle = -90) +
  scale_y_continuous(limits = c(0, 1))
```


* syntax
* aes()
* important types
* onedimensional
* barcounts
* density
* twodimensional
* line
* smoothed line
* ?
  * by groups
* colour
* facets

* pattern of gender differences by age group
* why are m and w more or less equal overall?
  * -> large group of old women with low turnout?
  