--- 
title: "Web Scraping using R"
author: "Jakob Tures"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: webscraping-tures/webscraping-tures.github.io
---

# Introduction {-}

This is an introduction to web scraping using R, designed to accompany a seminar
held at the University of Potsdam in summer 2021.

## What is web scraping? {-}

Web scraping encompasses methods of data collection from online sources. This
introduction will focus on scraping websites, but other forms on online data, 
e.g. databases or files, are scrapable as well.

Instead of copying and pasting content from a browser window, in web scraping 
the process is automated through a programming language as R or Python. This is
less error prone than manual collection and enables us to collect large amounts
of data efficiently or regularly repeat the scraping process to update the data
without investing much time. In web scraping your time is invested in planning 
the scraping process, programming your software and analysing the retrieved 
data -- aka the fun stuff -- rather than in time consuming and repetitive 
manual work.

Web scraping enables us to access data, that we could not retrieve using 
traditional methods of data collection. Mainly because some types of data would
not exist without the internet. Think of consumer reviews or social media 
posts. In other cases, web scraping enables us to collect data much more 
efficiently compared to traditional methods. Imagine you want to create a data 
file with demographic information in parliament members. This was obviously 
possible and done before web scraping became a possibility but included a lot of
manual work using printed almanacs and typing the information into a table. Time
consuming and error prone work. With web scraping we can set up scripts that 
gather this data directly from the website of a parliament including the 
possibility to  rerun the collection process after the next election. But we 
don't have to stop there. Maybe we want to analyse the sentiment of twitter 
messages written by parliament members by some socio-demographic variables as 
gender, age or educational background. We could retrieve both the tweets and the 
socio-demographic information using web scraping and analyse the gathered data
in combination.

At this point, your alarm bells should hopefully be ringing. Collecting data 
from social media profiles sounds like an Orwellian nightmare, and if it is done
incorrectly, it is. We should not and will not collect data for the sake of data
collection. Before scraping data on the internet we have to think about what 
data we actually need to answer our specific research question. Next we have to
check if we are actually allowed to collect certain types of data and even if we
are, if we should collect it. Do we need the names of people or users if we are
collecting data? Do we actually need all socio-demographic variables available 
or can we keep it at a minimum, ensuring that individuals in our data remain
unidentifiable? We are not web spiders, collecting every bit of data that is
available, whether we need it or not. We are scientists and have to behave in a
responsible and reasonable manner when collecting data.


## Course contents {-}

The first part of the course covers HTML as the conerstone of the (traditional) 
internet. We have to understand the structure of websites before we can start
picking out their specific parts. 

Part II serves as an introduction to R, the programming language we will use
for web scraping as well as analysis of the data gathered. This introduction can
necessarily only cover the basics of R, as our focus lies on introducing the
concepts of web scraping. The introduction to R will be problem-oriented, i.e. 
we learn what we need for web scraping, but there will be a lot left out or only
covered cursorily. See this as a starting point for your R journey, not its
destination. Additional resources for deepening your R knowledge will be 
provided at later points.

Part III forms the main part of the course and introduces basic and intermediate
web scraping concepts. We will learn how to identify data of interest on HTML
pages, select the data using CSS selectors and extract it using the R package
rvest. We will also look at scraping tables, content generated by dynamic 
websites and scraping of multi-page websites.

Part IV will cover the practical application of the acquired scraping knowledge
to a small real world data analysis project. In preparation, we will take a 
closer look on web scraping ethics and good practice, already briefly discussed
above, and what we can do to make sure we scrape responsibly. We will also 
deepen our R knowledge by looking on data cleaning, data transformation and
graphical analysis in R using the tidyverse packages. The course will close with
a sample project, spanning the process from generating a research question over
finding appropriate data and scraping it to cleaning and analysing the data with
the goal of answering the formulated question.

The internet is an ever changing landscape. Where once every page we ecnountered
was a simple HTML document, now many modern pages are less simplistic. Content 
is generated dynamically for each individual user and displayed information is
not actually "present" in the HTML files but loaded through JavaScript or 
database connections. This also makes scraping harder. While still being
possible, this requires advanced techniques that go beyond the scope of this 
introduction. You will still receive some pointers for where to look, should you
want to follow these paths further. Also, when we begin scraping raw text from
websites, we will need to learn about advanced handling of strings and regular 
expressions, whichcan also not be covered here. 

This introduction focusses on using rvest for scraping in R. This will suffice 
for many intermediate and advanced scraping projects, but alternatives exist
within R and in other languages. While these also can not be covered here, you
will gain a solid basis to pursue alternatives, should you so desire.


## Technical requirements {-}

All scraping in this introduction is done in R and RStudio using the tidyverse
packages. Information on how to install the software and handle packages will be
provided in chapters \@ref(R1) and \@ref(R2).

We will also need a browser for scraping. While in principle any popular browser
should do, I would recommend using Chrome or Chromium, as all examples using the
build in Web Developer Tools, are done in Chromium. You should make sure, that
the URL bar in your browser shows the full URLs. In Chrome/Chromium this is done
by right-clicking the URL bar and choosing "Always show full URLs" in the 
context menu.

While not strictly necessary, i would also advice using a decent text editor for
opening HTML files. The options that come with your operating systems but why 
not do yourselves a favour and use something you can actually work with. Options
are aplenty, but some of the more accessible while still powerful text editors I
personally would recommend are Notepad++ on Windows 
(<https://notepad-plus-plus.org/>{target="_blank"}) and Atom on all operating
systems (<https://atom.io/>{target="_blank"}). If your feel you are up to it, 
you can also dip your feet into Emacs, also available for every OS 
(<https://www.gnu.org/software/emacs/>{target="_blanK"}), but note that this is 
an advanced option. Proceed at your own risk.


## Conventions {-}

All code on the website is set in this `code font`.  This is true for R as well
as HTML code. R code that is included in text paragraphs will not always be 
runnable and often serve illustrative purposes. All R code written in code
blocks is runnable and look like this:

``` {r, collapse = TRUE}
print("Hello World!")
```

If an R command produces relevant output, you will see the output following 
directly after the command, written in red and introduced by `##`.

You should in general try to run the code blocks yourself in RStudio. The only 
way to learn what is happening in R and web scraping, is doing it yourself.
Run the examples and experiment with them. Every problem you will run into while
changing the code is a learning opportunity. Notice, that code blocks are
copyable. Mouse over the code block and you will see a copy symbol, or simply 
copy & paste like you are used to. But, I would strongly advice on writing as
much code as you can yourself. While copying code examples by hand may not be 
intellectually challenging, but if your aim is to learn the R functions and 
syntax you have to write a lot of it before it can become second nature.


## Acknowledgements {-}
I would like to thank the authors of bookdown the tidyverse and all R packages
used in the creation of this course on web scraping.

Very special thanks go to Lukas Höttges for his support in creating the website
and in organizing the seminar at the University of Potsdam. To Hannah Gehrmann
for the invaluable feedback on an early draft. To Fabian Class for technical 
support and numerous answers to my many questions. And to Sophia Varma for
translating the original German draft and the great support.

## Colophon {-}
``` {r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(knitr)
library(tidyverse)
library(rvest)
library(lubridate)
library(echarts4r)
```

The website was built with:

``` {r}
sessioninfo::session_info()
```
