--- 
title: "Web Scraping using R"
author: "Jakob Tures"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: webscraping-tures/webscraping-tures.github.io
---

# Introduction {-}

WIP!

## What is web scraping? {-}

* Goals
  * Accessing data that we cannot get with traditional methods.
  * Accessing data that is only conceivable on the internet 
  (i.e., customer reviews, user behaviour et al.)
  * What is feasible: Examples

* Methods
  * Identification of web content relevant to our query
  * Accessing web content directly from a programming language (R, Python, etc.)
  * Extraction of the page’s contents relevant to our interest
  * Date cleaning and transferral into a format suited to data analysis
  * Data analysis

## Course contents {-}

* Basic structure of the internet
  * HTML
  * Overview of other components we might encounter in the wild, without going too much in-depth

* Introduction to R
  * Basics in R Studio
  * Tidyverse
  * Basics in data transformation and graphical analysis

* Scraping
  * Scraping with rvest
  * CSS Selectors for selecting specific parts of a website
  * Scraping tables
  * Scraping dynamically generated websites (HTML forms)
  * Scraping multi-page websites
  * Good practice

* Not covered:
  * Handling of dynamic websites, JavaScript, databank access etc.
  * Handling of stings & regular expressions (Note: you will not be able to get around this if you want to follow this path)
  * Scraping with other packages or languages than R + rvest

## Technical requirements {-}

* R + RStudio
  * required packages in chap. 3

* A good text editor
  * Windows: Notepad ++
  * For all OS: Atom

* Browser
  * Chrome/Chromium recommended
  * Necessary settings (full URLs, WDTs)
  
## Conventions {-}

* Inline code and Code blocks
  * Also HTML code

* Code is copyable but use sparely (writing is important)

## Acknowledgements {-}
I would like to thank the authors of bookdown the tidyverse and all R packages
used in the creation of this course on web scraping.

Very special thanks go to Lukas Höttges for his support in creating the website
and in organizing the seminar at the University of Potsdam. To Hannah Gehrmann
for the invaluable feedback on an early draft. To Fabian Class for technical 
support and numerous answers to my many questions. And to Sophia Varma for
translating the original German draft and the great support.

## Colophon {-}
``` {r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(knitr)
library(tidyverse)
library(rvest)
library(lubridate)
library(echarts4r)
```

``` {r}
sessioninfo::session_info()
```
